## Road Casualties in Great Britain

Monthly totals of car drivers in Great Britain killed or seriously injured
January 1969 to December 1984. The compulsory wearing of seat belts was
introducted on January 31, 1983. These data were analyzed in @HarveyDurbin1986,
and used as an example in @DurbinKoopman2012, @Harvey1993a, and
@CommandeurKoopman2007.

This data is included in R with the dataset `Seatbelts`.

```{r Seatbelts}
data("Seatbelts", package = "datasets")
Seatbelts_ <- tbl_df(as.data.frame(Seatbelts)) %>%
  mutate(date = ts_to_date(time(Seatbelts))) %>%
  select(date, everything())
```

```{r Seatbelts-plot}
ggplot(Seatbelts_, aes(x = date, y = drivers)) +
  geom_point() +
  geom_line() +
  scale_y_log10() +
  xlab("")
```


```{r Seatbelts-StructTS-BSM}
StructTS(log10(Seatbelts[ , "drivers"]), type = "BSM")
```

```{r Seatbelts-Stan-BSM}
seatbelts_data <- within(list(), {
  y <- matrix(log10(Seatbelts_$drivers))
  n <- length(y)
  p <- 1
  q <- 3
  m <- 13
  T_lvls <- as.numeric(upper.tri(matrix(0, 2, 2), diag = TRUE))
  T_seas <- rbind(rep(-1, 11), diag(1, 10, 11))
  T <- matrix(0, 13, 13)
  T[1:2, 1:2] <- T_lvls
  T[3:13, 3:13] <- T_seas
  Z <- matrix(c(c(1, 0, 1), rep(0, 10)), 1, 13)
  R <- rbind(matrix(c(1, 0, 0,
                      0, 1, 0,
                      0, 0, 1), 3, 3),
             matrix(0, 10, 3))
  d <- array(0, 1)
  c <- rep(0, m)
  a1 <- rep(0, m)
  P1 <- diag(1e6, m, m) * var(as.numeric(y))
})
seatbelts_init <- list(sigma_eta = c(6.02, 1e-6, .02),
                      sigma_epsilon = .04)
seatbelts_bsm_sample <- sampling(var_only_mod,
                         chains = 1,
                         iter = 500,
                         init = list(seatbelts_init),
                         data = seatbelts_data, verbose = TRUE)
summary(seatbelts_bsm_sample, c("sigma_eta", "sigma_epsilon"))[[1]]
```



```{r Seatbelts-Stan-bsm-plot}
seatbelt_bsm_params <-
  extract_param_from_ssm_filter(rstan::extract(seatbelts_bsm_sample, "filter_res")[[1]], 13, 1)

seatbelt_bsm_states <- seatbelt_bsm_params[["a"]] %>%
  apply(2:3, mean) %>%
  reshape2::melt(varnames = c("obs", "state"))

ggplot(dplyr::filter(seatbelts_states, state <= 3),
       aes(x = obs, y = value)) +
  geom_line() +
  facet_wrap(~state, ncol = 1)
```


```{r Seatbelts-Stan-bsm-optimize}
seatbelts_optim <-
  rstan::optimizing(var_only_mod,
                   init = seatbelts_init,
                   data = seatbelts_data,
                   as_vector = FALSE,
                   verbose = TRUE)
seatbelts_optim$par[c("sigma_eta", "sigma_epsilon")]
```
