## CAPM

This is a simplified example of CAPM model from @Petris2010a.
A simplified capital-asset pricicing model (CAPM) model is a model of the excess
return (return relative to a risk-free asset) of stocks. In this example,
the excess return of a stock is proportional to the market return (return of a portfolio of stocks representing the whole market).


Let $y_{j,t}$ be the excess return of stock $j$ in period $t$.
The model that we will estimate is,
$$
\begin{aligned}[t]
\vec{y}_{j,t} &= \beta_{j,t} x_t + \varepsilon_{j,t} & \vec{\varepsilon}_t & \sim N(\vec{0}, \mat{\Sigma}_{\varepsilon} \\
\beta_{j,t} &= \beta_{j,t} + \eta_{j,t} & \vec{\eta}_{t} & \sim N(\vec{0}, \mat{\Sigma}_{\eta})
\end{aligned}
$$
The proportionality constants of the stocks are stock- and time-specific, with
each $\vec{\beta}_j$ following a random walk.[^alpha]
However, both the observation disturbances ($\mat{\Sigma}_{\varepsilon}$ and state disturbances ($\mat{\Sigma}_{\eta}$) allow for correlation between the stocks.
In the model estimated here, the intercepts for each stock are assumed to be zero.

The data, originally from @Berndt1991a, are the monthly returns of four stocks (Mobil, IBM, Weyer, and Citicorp) from January 1978 to December 1987.
The interest rate on 30-day Treasury Bill is used as the risk free rate (`rkfree`), and the value-weighted average returns of all stocks listed on the New York and American Stock Exchanges as the market return (`market`).
```{r capm-data}
data("capm", package = "StanStateSpace")
glimpse(capm)
```
In the analysis, the excess returns (return minus risk-free return) are used.
So, the first step is to subtract the risk-free returns (`rkfree`) from each of the stock return columns and the market return.
```{r capm-data-clean}
capm <- capm %>%
  mutate_each(funs(. - rkfree), one_of(c("mobil", "ibm", "weyer", "citcrp",
                                         "market"))) %>%
  mutate(obs = row_number())
ggplot(gather(select(capm, -rkfree), asset, rate, -date, -obs),
       aes(x = date, y = rate)) +
  geom_line() +
  facet_grid(asset ~ .) +
  theme_minimal()
```


```{r capm-model, results='hide', message = FALSE}
capm_mod <- ssm_stan_model("var_only_mv.stan")
```
```{r}
capm_data <-
  within(list(), {
    y <- as.matrix(capm[ , c("mobil", "ibm", "weyer", "citcrp")])
    n <- nrow(capm)
    p <- 4
    m <- p
    q <- m
    c_sz <- 1
    c <- array(0, c(c_sz, p))
    Z_sz <- n
    Z <- map(capm[["market"]], ~ diag(.x, p, p))
    d_sz <- 1
    d <- array(0, c(d_sz, m))
    T_sz <- 1
    T <- list(diag(1, m, m))
    R_sz <- 1
    R <- list(diag(1, m, q))
    a1 <- rep(1, m)
    P1 <- diag(1e3, m, m)
    # priors
    Rho_eta_prior <- 2
    Rho_epsilon_prior <- 2
    tau_epsilon_prior <- apply(y, 2, sd) * 0.5;
    tau_eta_prior <- rep(1, p)
  })

capm_samples <- sampling(capm_mod, chains = 1, iter = 1000,
                         data = capm_data)
capm_summary <- tidy_stan_summary(summary(capm_samples))[["all"]] %>%
  left_join(mutate(capm, obs = row_number()) %>% select(date, obs),
            by = c("dim_1" = "obs"))
```

```{r capm-state-plot}
capm_alpha <- filter(capm_summary, parameter == "alpha", dim_1 > 1) %>%
         mutate(stock = recode(dim_2,
                                `1` = "mobil",
                                `2` = "ibm",
                                `3` = "weyer",
                                `4` = "citcrp")) %>%
  left_join(gather(capm, stock, rate, -market, -date, -obs),
            by = c("date", "stock"))

capm_alpha_mean <- filter(capm_summary, parameter == "alpha_mean", dim_1 > 1) %>%
         mutate(stock = recode(dim_2,
                                `1` = "mobil",
                                `2` = "ibm",
                                `3` = "weyer",
                                `4` = "citcrp")) %>%
  left_join(gather(capm, stock, rate, -market, -date, -obs),
            by = c("date", "stock"))

ggplot(capm_alpha_mean,
       aes(x = date, color = stock, y = mean)) +
  geom_line() +
  theme_minimal()

ggplot(capm_alpha,
       aes(x = date,
           ymin = p97.5,
           ymax = p2.5)) +
  geom_ribbon(alpha = 0.3) +
  geom_line(aes(y = mean)) +
  facet_grid(stock ~ .) +
  theme_minimal()

```

```{r}
filtered <- 
  ssm_extract_summary(summary(capm_samples), 
                      "filtered", capm_data$m, capm_data$m, type = "filter")

capm_alphas_pred <- filter(filtered, parameter == "a", dim_1 > 1) %>%
         mutate(stock = recode(dim_2,
                                `1` = "mobil",
                                `2` = "ibm",
                                `3` = "weyer",
                                `4` = "citcrp")) %>%
  left_join(gather(capm, stock, rate, -market, -date, -obs),
            by = c("date", "stock"))

ggplot(capm_alphas_pred,
       aes(x = date,
           ymin = p2.5,
           ymax = p97.5)) +
  geom_ribbon(alpha = 0.3) +
  geom_line(aes(y = mean)) +
  facet_grid(stock ~ .) +
  theme_minimal()

ggplot(capm_alphas_pred, aes(x = date, y = mean)) +
  geom_line() +
  theme_minimal()

```

```{r capm-rho_epsilon_plot}
library("viridis")
ggplot(filter(capm_summary, parameter == "Rho_epsilon"),
        aes(fill = mean, x = dim_1, y = dim_2)) +
  geom_raster() +
  scale_fill_viridis()
```

```{r capm-Rho_eta_plot}
ggplot(filter(capm_summary, parameter == "Rho_eta"),
        aes(fill = mean, x = dim_1, y = dim_2)) +
  geom_raster() +
  scale_fill_viridis()
```

