# Example Models

## Nile

This is a short ($n = 100$) univariate time series of the annual flow volumes of
the Nile River at Aswan between 1871 and 1970. 
This series is described in @DurbinKoopman2012 and had been analyzed by @Cobb1978 and @Balke1993, and in numerous time series textbooks.
A notable feature of the series is a seeming structural break in 1899, around the time of the completion of the Aswan dam.

```{r Nile}
data("Nile", package = "datasets")
Nile_ <- data_frame(year = year(as.Date(Nile)),
                    flow = as.numeric(Nile))
```

```{r Nile_plot}
ggplot(Nile_, aes(x = year, y = flow)) +
  geom_point() +
  geom_line() +
  ylab("Annual Flow") + xlab("")
```

### Local Level Model

The Nile data can be modeled as a local level model,
$$
\begin{aligned}[t]
y_t &= \mu_t + \varepsilon_t & \varepsilon_t & \sim N(0, \sigma_{\varepsilon}^2) \\
\mu_{t + 1} &= \mu_t + \eta_t &
\eta_t & \sim N(0, \sigma^2_{\eta})
\end{aligned}
$$

```{r engine = 'cat', file = system.file("stan/models/local_level.stan", package = "StanStateSpace"), lang = 'stan'}
```

```{r Nile-Model, message=FALSE, warning=FALSE, eval=FALSE}
STAN_SSM_MODELS <- system.file("stan/models", package = "StanStateSpace")
STAN_SSM_INCLUDE <- system.file("stan/include", package = "StanStateSpace")
local_level_code <-
  stanc_builder(file.path(STAN_SSM_MODELS, "local_level.stan"),
                isystem = STAN_SSM_INCLUDE)
local_level_mod <- stan_model(model_code = local_level_code[["model_code"]],
                              model_name = local_level_code[["model_name"]])
```

```{r Nile-Local-Level-Sampling, eval = FALSE}
nile_data <- within(list(), {
  y <- matrix(Nile_$flow)
  n <- nrow(y)
  a1 <- array(0, 1)
  P1 <- matrix(10 ^ 7)
  y_scale <- sd(Nile_$flow)
})
nile_ll_sample <- sampling(local_level_mod,
                          chains = 4,
                          iter = 1000,
                          data = nile_data)
```
