## Nile

This is a short ($n = 100$) univariate time series of the annual flow volumes of
the Nile River at Aswan between 1871 and 1970.
This series is described in @DurbinKoopman2012 and had been analyzed by @Cobb1978 and @Balke1993, and in numerous time series textbooks.
A notable feature of the series is a seeming structural break in 1899, around the time of the completion of the Aswan dam.

```{r Nile}
data("Nile", package = "datasets")
Nile_ <- data_frame(year = year(as.Date(Nile)),
                    flow = as.numeric(Nile),
                    obs = seq_along(Nile))
```

```{r Nile_plot}
ggplot(Nile_, aes(x = year, y = flow)) +
  geom_point() +
  geom_line() +
  ylab("Annual Flow") + xlab("")
```

### Local Level Model

The Nile data can be modeled as a local level model,
$$
\begin{aligned}[t]
y_t &= \mu_t + \varepsilon_t & \varepsilon_t & \sim N(0, \sigma_{\varepsilon}^2) \\
\mu_{t + 1} &= \mu_t + \eta_t &
\eta_t & \sim N(0, \sigma^2_{\eta})
\end{aligned}
$$

```{r local_level_stan_code, echo = FALSE, comment = ""}
cat(paste(readLines(system.file("stan/models/local_level.stan", package = "StanStateSpace")), collapse = "\n"))
```

```{r Nile-Model, message=FALSE, warning=FALSE, results = 'hide'}
local_level_mod <- ssm_stan_model("local_level.stan")
```

```{r Nile-Local-Level-Sampling, message = FALSE}
nile_1_data <- within(list(), {
  y <- matrix(Nile_$flow)
  n <- nrow(y)
  a1 <- array(0, 1)
  P1 <- matrix(10 ^ 7)
  y_scale <- sd(Nile_$flow)
})
nile_1_samples <-
  sampling(local_level_mod,
           chains = 1,
           iter = 500,
           data = nile_1_data)

```

Now, summarize the MCMC samples using the `summary` function on the `stanfit` object.
Additionally, I use the `tidy_stan_summary` function to make the results of `summary` easier to work with.
This converts the results of `summary` from a list of matrices to a list of data frames, and also parses the parameter names so that it is easier to select particular parameter values by name. I also will only use only the summary statistics for the combined chains.
```{r nile_1_summary}
nile_1_summary <- tidy_stan_summary(summary(nile_1_samples))[["all"]] %>%
  left_join(Nile_, by = c("dim_1" = "obs"))
```
The estimated variances of the observation and state variances,
```{r summary_nile_1_samples_HQ}
filter(nile_1_summary, parameter %in% c("H", "Q")) %>%
  select(parname, mean, se_mean, p2.5, p97.5, n_eff, Rhat)
```
are similar to the MLE estimates producted by `StructTS`,
```{r nile_1_StructTS}
StructTS(Nile_$flow, type = "level")
```
However, since the Bayesian estimates are means, the MLE estimates are modes,
and the posterior distribution of the variances are right skewed, the means are
larger than the posterior modes.


```{r nile_1_states}
str_keep <- function(string, pattern) {
  string[str_detect(string, pattern)]
}

ggplot(filter(nile_1_summary, parameter == "alpha"),
       aes(x = year,
           ymin = mean - 2 * sd,
           ymax = mean + 2 * sd)) +
  geom_ribbon(alpha = 0.3) +
  geom_line(aes(y = mean)) +
  geom_point(aes(y = flow)) +
  ylab("Annual river flow") +
  xlab("Observation") +
  theme_minimal()

```



**TODO** Diagnostics. What are the relevant Bayesian analogs?


### Local level with known intervention (intercept)

```{r Nile-2-mod, results = 'hide', message = FALSE}
nile_2_mod <- ssm_stan_model("local_level_reg.stan")
```

```{r Nile-2-run}
nile_2_data <- nile_1_data
nile_2_data[["x"]] <- matrix(as.integer(Nile_$year > 1899))
nile_2_data[["k"]] <- ncol(nile_2_data[["x"]])
nile_2_samples <- sampling(nile_2_mod, chains = 1, iter = 500,
                           data = nile_2_data)
```

```{r Nile-2-summary}
nile_2_summary <- tidy_stan_summary(summary(nile_2_samples))[["all"]] %>%
  left_join(Nile_, by = c("dim_1" = "obs"))
```


```{r nile_2_states}
ggplot(filter(nile_2_summary, parameter == "mu"),
       aes(x = year,
           ymin = mean - 2 * sd,
           ymax = mean + 2 * sd)) +
  geom_ribbon(alpha = 0.3) +
  geom_line(aes(y = mean)) +
  geom_point(aes(y = flow)) +
  ylab("Annual river flow") +
  xlab("Observation") +
  theme_minimal()

```


### Local Level with known intervention (variance)

```{r Nile-local-level-interven, results = 'hide', message = FALSE}
nile_3_mod <- ssm_stan_model("local_level_interven.stan")
```

```{r Nile-3-run}
nile_3_data <- nile_1_data
nile_3_data[["s"]] <- ifelse(Nile_$year == 1899, 10, 1)
nile_3_samples <- sampling(nile_3_mod, chains = 1, iter = 500,
                           data = nile_3_data)
```

```{r Nile-3-summary}
nile_3_summary <- tidy_stan_summary(summary(nile_3_samples))[["all"]] %>%
  left_join(Nile_, by = c("dim_1" = "obs"))
```


```{r nile_3_states}
ggplot(filter(nile_3_summary, parameter == "alpha"),
       aes(x = year,
           ymin = mean - 2 * sd,
           ymax = mean + 2 * sd)) +
  geom_ribbon(alpha = 0.3) +
  geom_line(aes(y = mean)) +
  geom_point(aes(y = flow)) +
  ylab("Annual river flow") +
  xlab("Observation") +
  theme_minimal()

```

### Local Level model with Sparse State Disturbances

```{r Nile-local-level-tvvar, results = 'hide', message = FALSE}
nile_4_mod <- ssm_stan_model("local_level_tvvar.stan")
```

```{r Nile-4-run}
nile_4_data <- nile_1_data
nile_4_data[["s"]] <- 1 / nrow(Nile_)
nile_4_samples <- sampling(nile_4_mod, chains = 1, iter = 500,
                           data = nile_4_data)

```

```{r Nile-4-summary}
nile_4_summary <- tidy_stan_summary(summary(nile_4_samples))[["all"]] %>%
  left_join(Nile_, by = c("dim_1" = "obs"))
```


```{r nile_4_states}
ggplot(filter(nile_4_summary, parameter == "alpha"),
       aes(x = year,
           ymin = mean - 2 * sd,
           ymax = mean + 2 * sd)) +
  geom_ribbon(alpha = 0.3) +
  geom_line(aes(y = mean)) +
  geom_point(aes(y = flow)) +
  ylab("Annual river flow") +
  xlab("Observation") +
  theme_minimal()
```
